---
title: "preprocessing"
author: "Maisy Hallam"
date: "2026-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preprocessing experiment data

First, import all participants' data, concatenate all files, and remove all identifiable information from the data (i.e. Prolific IDs).

(NOTE: The data available on OSF has already been concatenated and anonymised. You do not have access to the raw, non-anonymous data, so you cannot run the following chunk.)

```{r}
base_dir <- 'data/raw/participants/' # a directory that contains a CSV file for each participant

all_files <- paste0(base_dir, list.files(base_dir, recursive = TRUE))

ppt_data <- read_csv(all_files) %>% # read in all files from this directory
  filter(random_id != 'random_id') %>% # get rid of duplicate headers, i.e. where the value of 'random_id' is 'random_id'
  select(-c(prolific_id)) # and remove prolific IDs
```

The anonymised `ppt_data` dataframe is saved to our `raw` data directory.

(If you've downloaded from OSF, this is the only raw data you have available.)

```{r}
write_csv(ppt_data, 'data/raw/anonymised_data.csv')
```


Next, before we analyse this data, we want to remove data from participants who failed an attention check.

We also want to remove all rows that contain attention check data, since we won't need those for analysis.

```{r}
failed <- filter(ppt_data,feature1 == 'attn' & correct == 'N')$random_id

clean_data <- ppt_data %>% 
  filter(!random_id %in% failed & feature1 != 'attn') # remove the data of anyone who failed the attention check, also remove all attention check data in general
```

Now we can save this new dataframe `clean_data` to our `preprocessed` data directory.

```{r}
write_csv(clean_data,'data/preprocessed/experiment_data.csv')
```

