---
title: "preprocessing"
author: "Maisy Hallam"
date: "2026-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Preprocessing data for kin term granularity analyses

In this document, we will preprocess the cross-linguistic data on kin term granularity compiled by our model, and extract data on population size from the LinguaMeta dataset and combine it with the kinship data.

We will also extract data from D-PLACE on four variables: settlement (EA030), subsistence (EA042), household size (B022), and family size (B023) (Appendix).

## Kin term granularity

First, let's import the data from our typological analysis. `gran` contains six columns:
* `language` is the language name
* `language_family` is the language family
* `glottocode` is the glottocode for that language
* `prop` is the kin term granularity: the proportion of kin terms over kin types.
* `count` is the raw number of kin terms in that language.
* `n_kintypes` is the number of kin types that were available in that language's dataset (out of a possible 48).

```{r}
gran <- read_csv('data/raw/granularity.csv')

gran
```

First, we will filter this dataset so it only contains languages who have at least 40/48 kin types in their dataset (`n_kintypes` > 40), to preserve the accuracy of the kin term granularity measure as much as possible.

```{r}
gran_filtered <- gran %>% filter(n_kintypes > 40)

write_csv(gran_filtered,'data/preprocessed/kinterm_granularity.csv')
```

## Population size

Our population size data come from LinguaMeta, so we import the full LinguaMeta dataset and filter just the `estimated_number_of_speakers` variable.

```{r}
pop_size <- read_tsv('data/raw/linguameta.tsv') %>% select(c(english_name,estimated_number_of_speakers,glottocode))
```

Then we merge this data with the `lex_size_filtered` data:

```{r}
gran_pop <- merge(gran_filtered,pop_size, .by = 'glottocode', all.x = TRUE) %>% filter(!is.na(estimated_number_of_speakers)) %>% # remove NA values
  dplyr::select(c(glottocode,language,estimated_number_of_speakers,family,prop,n_kintypes)) # select only relevant columns

gran_pop
```

We will want to take the log estimated number of speakers for our analyses, so we create a new variable `log_pop`.

```{r}
gran_pop <- gran_pop %>% mutate(
  log_pop = log2(estimated_number_of_speakers)
)
```

And save this new dataframe.

```{r}
write_csv(gran_pop,'data/preprocessed/granularity_population.csv')
```


## D-PLACE data

We compare kin term granularity to some other correlates of network density: settlement type, subsistence type, average household size, and average family size.

First, import the data from D-PLACE:

It's split into three dataframes: 
* `codes` contains each variable in D-PLACE and its possible values.
* `data` contains the value for each variable for each society (by ID)
* `societies` contains each society and their corresponding ID.

```{r}
codes <- read_csv('data/raw/dplace/codes.csv')
data <- read_csv('data/raw/dplace/data.csv')
societies <- read_csv('data/raw/dplace/societies.csv')
```

We need to merge the dataframes together in order to have both the data and the society name in a single dataframe.

Notice the column `ID` in `societies` does not match the column name `Soc_ID` in `data`; let's fix that.

```{r}
colnames(societies)
colnames(data)

societies_renamed <- societies %>% rename(Soc_ID = ID)
```

Now let's merge `societies` and `data` by the `Soc_ID`.

```{r}
societies_data <- merge(societies_renamed,data,by = 'Soc_ID')
```

And then filter by our first variable, EA030 (settlement type).
We'll also select only a handful of columns: we want the society `Name`,
the language's `Glottocode`, and the `Value` for the settlement type variable.

```{r}
settlement <- filter(societies_data,Var_ID == 'EA030') 

%>% dplyr::select(c(Name,Glottocode,Value))

settlement
```

We have our own coding system that is less granular than the D-PLACE system, so we'll create a new column with our own Values.

```{r}
settlement <- settlement %>% mutate(
    Settlement_Type = case_when(
      Value %in% c('Nomadic','Seminomadic','Semisedentary','Impermanent') ~ 'Nomadic/Impermanent',
      Value %in% c('Dispersed homesteads','Hamlets') ~ 'Small settlements',
      Value %in% c('Villages/towns','Complex permanent') ~ 'Large settlements'
    )
  )
```

And save this new dataframe.

```{r}
write_csv(settlement,'data/preprocessed/settlement_type.csv')
```


Now we'll do the same for our subsistence type data.

```{r}
subsistence <- filter(societies_data,Var_ID == 'EA042') %>% dplyr::select(c(Name,Glottocode,Value))

subsistence <- subsistence %>% mutate(
    Subsistence_Type = case_when(
      Value %in% c('Gathering','Fishing','Hunting','Pastoralism') ~ 'Hunting and gathering',
      Value %in% c('Casual agriculture','Extensive agriculture', 'Intensive agriculture', 'Agriculture, type unknown') ~ 'Agriculture')
    )

write_csv(subsistence,'data/preprocessed/subsistence_type.csv')
```

And for household size data:

```{r}
household <- societies_data %>% filter(Var_ID == 'B022') %>% 
  dplyr::select(Name, Glottocode, Value) %>%
  rename(Household_Size = Value)

household

write_csv(household, 'data/preprocessed/household_size.csv')
```

And family size data:

```{r}
family <- societies_data %>% filter(Var_ID == 'B023') %>% 
  dplyr::select(Name, Glottocode, Value) %>%
  rename(Family_Size = Value)

family

write_csv(family, 'data/preprocessed/family_size.csv')
```

