---
title: "preprocessing"
author: "Maisy Hallam"
date: "2026-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Preprocessing data for kin term granularity analyses

In this document, we will preprocess the cross-linguistic data on kin term granularity compiled by our model (Section 2 and 3 of the paper)

## Kin term granularity

First, let's import the data from our typological analysis. `lex_size` contains six columns:
* `language` is the language name
* `language_family` is the language family
* `glottocode` is the glottocode for that language
* `prop` is the kin term granularity: the proportion of kin terms over kin types.
* `count` is the raw number of kin terms in that language.
* `n_kintypes` is the number of kin types that were available in that language's dataset (out of a possible 48).

```{r}
gran <- read_csv('data/raw/granularity.csv')

```

First, we will filter this dataset so it only contains languages who have at least 40/48 kin types in their dataset (`n_kintypes` > 40), to preserve the accuracy of the kin term granularity measure as much as possible.

```{r}
gran_filtered <- gran %>% filter(n_kintypes > 40)
```

We'll be comparing this data against data produced by a model, so we want to add a column that marks this data as veridical:

```{r}
gran_filtered <- gran_filtered %>% mutate(data_type = 'veridical')
```

And save.

```{r}
write_csv(gran_filtered,'data/preprocessed/kinterm_granularity.csv')
```

## Model data

The baseline data has the same columns as the natural language data.

```{r}
baseline <- read_csv('data/raw/simulation.csv')
```

Again, we want to add a column to mark this as data from the model.

```{r}
baseline <- baseline %>% mutate(data_type = 'simulated')
```

And save.

```{r}
write_csv(baseline, 'data/preprocessed/simulated_baseline.csv')
```

