---
title: "social-practice-ordinal"
author: "Maisy Hallam"
date: "2026-01-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ordinal)
library(ggeffects)
library(emmeans)
library(patchwork)
```

```{r}
theme_set(theme_bw(base_size = 10))
theme_update(axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = 'black'),
    text = element_text(family = "Montserrat"),
    plot.margin = margin(c(0,0,0,0)),
    axis.text = element_text(size = 8),
    panel.background = element_rect(fill = 'transparent'),
    plot.background = element_rect(fill = 'transparent', color = 'NA'))
```


# Are kin term semantics predicted by social practice?

To model whether kin term semantics vary with social practice, here I will test whether the rank of a particular feature can be predicted by a social practice variable.

We test four predictions here:
1. Societies with bilateral descent will have higher ranks on Lineality.
2. Societies with unilineal descent will have higher ranks on GCR.
3. Societies who allow cousin marriage will have higher ranks on Lineality.
4. Societies who allow cross-cousin marriage only will have higher ranks on GCR.


First, we need to import the kin term semantics data and create a new variable, `rank`,
based on which features are weighted highest and which are weighted lowest.

The highest ranked feature receives rank 6, the lowest receives rank 1. 

```{r}
sem <- read_csv('data/preprocessed/model_run.csv')

sem_l <- sem %>% pivot_longer(cols = c(generation,gender,lineal,gcr,seniority,speaker_gender),names_to = 'feature',values_to = 'weight') %>% unique()

ranked <- sem_l %>% group_by(Name) %>% summarise(
  Glottocode = Glottocode,
  Family = Family,
  Feature = feature,
  Weight = weight,
  Rank = case_when(
    weight == sort(Weight)[1] ~ 1,
    weight == sort(Weight)[2] ~ 2,
    weight == sort(Weight)[3] ~ 3,
    weight == sort(Weight)[4] ~ 4,
    weight == sort(Weight)[5] ~ 5,
    weight == sort(Weight)[6] ~ 6
  )
)

ranked
```

Now we have a dataset that includes rankings, let's import our social practice data.

```{r}
descent <- read_csv('data/preprocessed/descent.csv') %>% rename(Glottocode = glottocode)

marriage <- read_csv('data/preprocessed/marriage.csv') %>% rename(Glottocode = glottocode)
```


# BILATERAL SOCIETIES RANK LINEALITY HIGHER

Combine `descent` data with `ranked` data, and create `Descent_Type` variable.

```{r}
descent_ranked <- merge(ranked,descent,.by = 'Glottocode') %>% 
  mutate(
  Descent_Type = factor(case_when(
    Value %in% c('Mixed','Bilateral', 'Ambilineal') ~ 'bilateral',
    Value %in% c('Matrilineal','Patrilineal') ~ 'unilineal',
  ))
) %>%
  dplyr::select(
    c(Glottocode, Name, Family, Feature, Rank, Descent_Type)
  )

```

Set factors.

```{r}
descent_ranked <- descent_ranked %>% mutate(
  Descent_Type = factor(Descent_Type, levels = c('unilineal', 'bilateral')), # unilineal is the reference level
  Family = factor(Family),
  Rank = factor(Rank, ordered = TRUE)
)

levels(descent_ranked$Descent_Type)
```

Filter data to only include Lineality rank, and then plot a histogram of
how many societies have each ranking.

```{r}
descent_lineality <- descent_ranked %>% filter(Feature == 'lineal')

descent_lineality %>%
  ggplot(aes(x = Rank, fill = Descent_Type)) +
  geom_histogram(stat = 'count')
```

Run the ordinal model predicting `Rank` of lineality by `Descent_Type` (with random intercepts for Language Family).

```{r}
dl_mdl <- clmm(Rank ~ Descent_Type + (1|Family), data = descent_lineality)

summary(dl_mdl)
```

The estimate for `bilateral` is 0.5447 log-odds, which means that when a society is bilateral, the odds of Lineality being ranked higher are approximately doubled:

```{r}
exp(0.5447)
```

Plotting the model predictions, we can that bilateral societies tend to have higher ranks on Lineality,
and unilineal societies tend to have lower ranks:

```{r}
pred <- ggpredict(dl_mdl, terms = "Descent_Type")

plot(pred)
```


# UNILINEAL SOCIETIES RANK GCR HIGHER

Combine `descent` data with `ranked` data.

```{r}
descent_ranked # same df as above
```

Set factors - this time, we want `bilateral` to be the reference level.

```{r}
descent_ranked <- descent_ranked %>% mutate(
  Descent_Type = factor(Descent_Type, levels = c('bilateral', 'unilineal'))) # bilateral is the reference level

levels(descent_ranked$Descent_Type)
```

Filter data (as we are only interested in the rank of GCR):

```{r}
descent_gcr <- descent_ranked %>% filter(Feature == 'gcr')

descent_gcr %>%
  ggplot(aes(x = Rank, fill = Descent_Type)) +
  geom_histogram(stat = 'count')
```

Run the ordinal model.

```{r}
dg_mdl <- clmm(Rank ~ Descent_Type + (1|Family), data = descent_gcr)

summary(dg_mdl)
```

The estimate for `unilineal` is 0.9068 log-odds, which means that when a society is unilineal, the odds of GCR being ranked higher are more than doubled:

```{r}
exp(0.9068)
```

Again this is confirmed visually by plotting the model predictions:

```{r}
pred <- ggpredict(dg_mdl, terms = "Descent_Type")

plot(pred)
```

# SOCEITIES WHO ALLOW COUSIN MARRIAGE RANK LINEALITY HIGHER

Combine `marriage` data with `ranked` data, and create two variables: `Marriage_Permitted`,
which tells us if a society permits marriage to any or all cousins, and `Marriage_Type` which tells us
which cousins are permitted (for the cross-cousin marriage - GCR analysis).

```{r}
cousin_marriage <- merge(ranked,marriage,.by = 'Glottocode') %>% 
  mutate(
  Marriage_Permitted = ifelse(Value %in% c('No first/second cousins', 'No first cousins'), 'N','Y'),
  Marriage_Type = factor(case_when(
    Value %in% c('Cross-cousin', 'Matrilateral cross only', 'Patrilateral cross only') ~ 'cross cousins',
    Value %in% c('Any first cousins', 'Paternal only', 'Maternal only') ~ 'other cousins',
    Value %in% c('No first/second cousins', 'No first cousins') ~ 'no cousins'
))) %>%
  dplyr::select(
    c(Glottocode, Name, Family, Feature, Rank, Marriage_Type,Marriage_Permitted)
  )

cousin_marriage
```

Set factors.

```{r}
cousin_marriage <- cousin_marriage %>% mutate(
  Marriage_Permitted = factor(Marriage_Permitted, levels = c('N','Y')), # N is the reference level
  Family = factor(Family),
  Rank = factor(Rank, ordered = TRUE)
)

levels(cousin_marriage$Marriage_Permitted)
```

Filter data (as we are only interested in the rank of Lineality), then make a histogram
of how often each rank is assigned when a society permits/does not permit cousin marriage.

```{r}
marriage_lineality <- cousin_marriage %>% filter(Feature == 'lineal')

marriage_lineality %>%
  ggplot(aes(x = Rank, fill = Marriage_Permitted)) +
  geom_histogram(stat = 'count')
```

Run the ordinal model.

```{r}
ml_mdl <- clmm(Rank ~ Marriage_Permitted + (1|Family), data = marriage_lineality)


summary(ml_mdl)
```

The estimate for `Yes` is 0.05438 log-odds, which means when a society permits cousin marriage the odds of Lineality being ranked higher are approximately the same as in a society where no cousin marriage is permitted:

```{r}
exp(0.05438)
```

And indeed the model predicts no difference between the probability of each rank:

```{r}
pred <- ggpredict(m_l_mdl2, terms = "Marriage_Permitted")

plot(pred)
```

# SOCIETIES WHO ALLOW CROSS-COUSIN MARRIAGE RANK GCR HIGHER

Combine `marriage` data with `ranked` data, and create a `Marriage_Type` variable which tells us what
kinds of cousins you are allowed to marry in that society.

```{r}
cousin_marriage # same df as above
```

Set factors.

```{r}
cousin_marriage <- cousin_marriage %>% mutate(
  Marriage_Type = factor(Marriage_Type, levels = c('no cousins', 'other cousins', 'cross cousins')), # no cousins is the reference level
  Family = factor(Family),
  Rank = factor(Rank, ordered = TRUE)
)

levels(cousin_marriage$Marriage_Type)
```

Filter data (as we are only interested in the rank of GCR):

```{r}
marriage_gcr <- cousin_marriage %>% filter(Feature == 'gcr')

marriage_gcr %>%
  ggplot(aes(x = Rank, fill = Marriage_Type)) +
  geom_histogram(stat = 'count')
```

Run the ordinal model.

```{r}
mg_mdl <- clmm(Rank ~ Marriage_Type + (1|Family), data = marriage_gcr)

summary(mg_mdl)
```

The estimate for `other cousins` is -0.8611 log-odds, which means that when a society allows cousin marriage (but NOT cross-cousin marriage specifically), the odds of GCR being ranked higher decrease by about half compared to a society that permits no cousin marriage. But this effect is not significant.

```{r}
exp(-0.8611)
```

The estimate for `cross cousins` is 0.1038 log-odds, which means that when a society allows marriage to cross cousins, the odds of GCR being ranked higher are about the same as a society that allows no cousin marriage.

```{r}
exp(0.1038)
```

Plotting the model predictions shows that the probability of each rank is about the same for each group,
but a rank of 1 seems a bit more likely for the `other cousins` group.

```{r}
pred <- ggpredict(mg_mdl, terms = "Marriage_Type")

plot(pred)
```

We'll use `emmeans` to look at the pairwise comparisons for this model, which shows that there are
no significant increases or decreases in rank of GCR between any of the groups.

```{r}
emm <- emmeans(mg_mdl, ~ Marriage_Type)
contrast(emm, "pairwise", type = "response")
```

# Plot the distribution of ranks

Finally, we'll plot this data tidily for the paper.
NOTE: if you do not have the font 'Charter' installed, please comment out the line that assigns fonts.

Set plot themes:

```{r}
extrafont::loadfonts()

theme_set(theme_bw(base_size = 10))
theme_update(axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = 'black'),
    text = element_text(family = "Charter"), # comment out this line if you do not have the font Charter
    panel.background = element_rect(fill = 'transparent'),
    plot.background = element_rect(fill = 'transparent', color = 'NA'))
```


First, a plot for Descent Type + Lineality rank.

```{r}
colours = c(
  'bilateral' = '#f6ecd0',
  'unilineal' = '#708090')


bl <- descent_lineality %>% filter(!is.na(Descent_Type)) %>% mutate(Descent_Type = factor(Descent_Type, levels = c('bilateral','unilineal'))) %>%
  ggplot(aes(x = fct_rev(Rank), fill = Descent_Type)) + # fct_rev reverses the order of factors
                                                        # for a more intuitive x axis
  scale_fill_manual(name = 'Descent Type', values = colours) +
  geom_bar(position = 'dodge', linewidth = 0.2, colour = 'black') +
  labs(x = 'Rank of Lineality',
       y = 'count',
       fill = 'Descent type',
       title = 'Descent type and Lineality') +
  theme(legend.position = 'none',
        plot.margin = margin(c(0,0,10,0)),
        plot.title = element_text(size = 7))

bl

descent_lineality %>% 
  filter(!is.na(Descent_Type)) %>% 
  mutate(Descent_Type = factor(Descent_Type, levels = c('bilateral','unilineal')),
         n = n()) %>%
  group_by(Descent_Type, Rank) %>%
  mutate(
    prop_n = n() / n
  ) %>%
ggplot(aes(fill = Descent_Type)) +
  scale_fill_manual(name = 'Descent Type', values = colours) +
  geom_point(aes(x = Rank), stat = 'count', shape = 23, size = 5)



```

A plot for Descent Type and GCR:

```{r}
ug <- descent_gcr %>% filter(!is.na(Descent_Type)) %>%
  ggplot(aes(x = fct_rev(Rank), fill = Descent_Type)) +
  scale_fill_manual(name = 'Descent Type', values = colours) +
  geom_bar(position = 'dodge', linewidth = 0.2, colour = 'black') +
  labs(x = 'Rank of GCR',
       y = '',
       fill = 'Descent type',
       title = 'Descent type and GCR') +
  theme(legend.margin = margin(c(1,1,1,1)),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7),
        legend.key.size = unit(0.4, 'cm'),
        plot.margin = margin(c(0,0,10,0)),
        plot.title = element_text(size = 7))

ug
```

Cousin marriage and Lineality.

```{r}
colours = c(
  'N' = '#f6ecd0',
  'Y' = '#708090')

ml <- marriage_lineality %>% filter(!is.na(Marriage_Permitted)) %>%
  ggplot(aes(x = fct_rev(Rank), fill = Marriage_Permitted)) +
  scale_fill_manual(name = 'Marriage Permitted', values = colours) +
  geom_bar(position = 'dodge', linewidth = 0.2, colour = 'black') +
  labs(x = 'Rank of Lineality',
       y = 'count',
       fill = 'Marriage permitted',
       title = 'Cousin Marriage and Lineality') +
  theme(legend.margin = margin(c(0,0,0,0)),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7),
        legend.key.size = unit(0.4, 'cm'),
        plot.margin = margin(c(0,0,0,0)),
        plot.title = element_text(size = 7))
```

Cousin marriage and GCR.

```{r}
colours = c(
  'no cousins' = '#f6ecd0',
  'other cousins' = '#666',
  'cross cousins' = '#708090')

mg <- marriage_gcr %>% filter(!is.na(Marriage_Type)) %>% mutate(Marriage_Type = factor(Marriage_Type, levels = c('no cousins', 'cross cousins', 'other cousins'))) %>%
  ggplot(aes(x = fct_rev(Rank), fill = Marriage_Type)) +
  scale_fill_manual(name = 'Marriage Permitted', values = colours) +
  geom_bar(position = 'dodge', linewidth = 0.2, colour = 'black') +
  labs(x = 'Rank of GCR',
       y = '',
       fill = 'Marriage type',
       title = 'Cross-cousin Marriage and GCR') +
  theme(legend.margin = margin(c(0,0,0,0)),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7),
        legend.key.size = unit(0.4, 'cm'),
        plot.margin = margin(c(0,0,0,0)),
        plot.title = element_text(size = 7))
```



Combine all four plots together using `patchwork` package, and save.

```{r}
bl + ug + ml + mg

ggsave('plots/social_practice.pdf', device = cairo_pdf, width = 14, height = 10, unit = 'cm')
```

