---
title: "preprocessing"
author: "Maisy Hallam"
date: "2026-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preprocessing data for social practice analysis

In this document, we will filter out variables EA023 (cousin marriage) and EA043 (patterns of descent)
from the full D-PLACE dataset.


First, import the data from D-PLACE:

It's split into three dataframes: 
* `codes` contains each variable in D-PLACE and its possible values.
* `data` contains the value for each variable for each society (by ID)
* `societies` contains each society and their corresponding ID.

```{r}
codes <- read_csv('data/raw/codes.csv')
data <- read_csv('data/raw/data.csv')
societies <- read_csv('data/raw/societies.csv')
```

We need to merge the dataframes together in order to have both the data and the society name in a single dataframe.

Notice the column `id` in `societies` does not match the column name `Soc_ID` in `data`; let's fix that.

```{r}
colnames(societies)
colnames(data)

societies_renamed <- societies %>% rename(Soc_ID = id)
```

Now let's merge `societies` and `data` by the `Soc_ID`.

```{r}
societies_data <- merge(societies_renamed,data,by = 'Soc_ID')
```

And then filter by our first variable, EA024 (cousin marriage practices).
We'll also select only a handful of columns: we want the society name `pref_name_for_society`,
the language's `glottocode`, the `Value` for the cousin marriage variable.

```{r}
marriage <- filter(societies_data,Var_ID == 'EA023') %>% dplyr::select(c(pref_name_for_society,glottocode,Value,pref_name_for_society))
```

And then filter by our second variable, EA043 (patterns of descent).
Again we only select a handful of columns.

```{r}
descent <- filter(societies_data,Var_ID == 'EA043') %>% dplyr::select(c(pref_name_for_society,glottocode,Value,pref_name_for_society))
```

Save these new dataframes to our `preprocessed` directory. NOTE: these dataframes already exist in this directory - please only run this cell if you are sure you want to rewrite them.

```{r}
write_csv(marriage,'data/preprocessed/marriage.csv')

write_csv(descent, 'data/preprocessed/descent.csv')
```

