---
title: "analysis"
author: "Maisy Hallam"
date: "2026-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggtext)
library(extrafont)
library(ggExtra)
library(patchwork)
library(lme4)
library(broom)
library(emmeans)
library(binom)
```

```{r}
# PLOT THEMES AND COLOURS

extrafont::loadfonts()

theme_set(theme_bw(base_size = 10))
theme_update(axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = 'black'),
    text = element_text(family = "Charter"),
    plot.margin = margin(c(0,0,0,0)),
    axis.text = element_text(size = 8),
    panel.background = element_rect(fill = 'transparent'),
    plot.background = element_rect(fill = 'transparent', color = 'NA'))
```

Import experiment data.

```{r}
# participant z scores compared to their bespoke baseline
ppt_z <- read_csv('data/by_participant_z.csv') %>% mutate(system = factor(system))

# participant 'accuracy' - whether their generalisations increased predictive structure
acc <- read_csv('data/accuracy.csv') %>% mutate(system = factor(system), random_id = factor(random_id), correct_01 = factor(correct))

# whether or not the participants encoded gender distinctions between gender distinct pairs of referents
gender <- read_csv('data/gender_distinctions.csv') %>% mutate(gender = factor(distinction))

# proportion with which each referent was labelled the same as each other referent
distance <- read_csv('data/participant_distance.csv')

perf_distance <- read_csv('data/ideal_distance.csv')
```

# Section 4: does predictive structure emerge during generalisation?

```{r}
acc %>% group_by(random_id,system) %>%
  summarise(prop_correct = sum(correct/n())) %>%
  filter(prop_correct == 1)
```

```{r}
ppt_z %>% filter(z < -1.96) %>% merge(acc) %>%
  group_by(random_id,system) %>%
  summarise(
    new_term = ifelse(chosen_term != gen1_term & chosen_term != gen2_term, 1, 0),
    n = n(),
    proportion = sum(new_term) / n()) %>%
  summarise(
    any_new_term = ifelse(proportion > 0, 1,0)) %>% unique() %>%
  group_by() %>%
  mutate(full_prop = sum(any_new_term) / 106)

```

```{r}
ppt_z %>% filter(z > -1.96) %>% merge(acc) %>%
  group_by(random_id,system) %>%
  summarise(
    new_term = ifelse(chosen_term != gen1_term & chosen_term != gen2_term, 1, 0),
    n = n(),
    proportion = sum(new_term) / n()) %>%
  summarise(
    any_new_term = ifelse(proportion > 0, 1,0)) %>% unique() %>%
  group_by() %>%
  mutate(full_prop = sum(any_new_term) / 179)
```


### Figure 11. Proportion of predictive generalisations made by participants across all eight trials.


Predictive generalisations across all trials (left-hand plot):

```{r}
all_accuracy_plt <- acc %>%
  
  group_by(random_id,system) %>%
  
  # get the proportion of 'correct' generalisations for each participant per system type
  summarise(
    prop_correct = sum(correct) / n(),
    n_trials = n(),
    system = factor(system),
    system = system) %>%
  
  unique() %>%
  
  group_by() %>%
  
  # get cross-participant means and confidence intervals for each system type
  mutate(
    mean_propn_expected = mean(prop_correct),
    sd = sd(prop_correct),
    n = n(),
    se = sd/sqrt(n),
    lower95 = mean_propn_expected - 1.96 * se,
    upper95 = mean_propn_expected + 1.96 * se
  ) %>%
  
  ggplot(aes(x = 1, y = prop_correct)) +
  geom_hline(yintercept = 0.25,linetype = 'dashed') +

  geom_violin(fill= '#D0453B', colour = '#D0453B', alpha = 0.8, trim = FALSE) +
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.05, linewidth = 0.01,colour = 'black', alpha = 0.7) +

  geom_point(aes(y = mean_propn_expected), alpha = 0.7, shape = 18) +
  labs(
       x = 'All participants',
       y = 'Proportion of \n predictive generalisations'
       ) +
  ylim(c(0,1)) +
  theme(legend.position = 'none',
        axis.text.x = element_blank())

all_accuracy_plt
```

Predictive generalisations by system type (left-hand plot):

```{r}
# group each system by which Type it belongs to for axis labels
acc <- acc %>%
  mutate(
    system_type = case_when(
      system == 0 ~ 'Type I',
      system %in% c(1,2,3) ~ 'Type II',
      system %in% c(4,5,6,7) ~ 'Type III'),
    system_type = factor(system_type, levels = c('Type I','Type II','Type III'))
    )

type_acc_plt <- acc %>%  
  group_by(random_id, system, system_type) %>%
  
  # get proportion per participant per system
  summarise(
    prop_correct = sum(correct) / n(),
    n_trials = n()) %>%
  
  group_by(system_type) %>%
  
  # get means and confidence intervals across participants (within system type)
  summarise(
    prop_correct = prop_correct,
    mean_propn_expected = mean(prop_correct),
    sd = sd(prop_correct),
    n = n(),
    se = sd/sqrt(n),
    lower95 = mean_propn_expected - 1.96 * se,
    upper95 = mean_propn_expected + 1.96 * se,
    system_group = factor(system_type, levels = c('Type I','Type II','Type III'))
  ) %>%
  
  ggplot(aes(x = system_type, y = prop_correct)) +
  scale_fill_manual(name = 'System type', labels = c('Type I','Type II','Type III'),values=c('#D0453B','#E96922','#66748F')) +
  scale_colour_manual(name = 'System type', labels = c('Type I','Type II','Type III'),values= c('#D0453B','#E96922','#66748F')) +
    geom_hline(yintercept = 0.25,linetype = 'dashed') +

  geom_violin(aes(fill= system_type, colour = system_type), alpha = 0.8, trim = FALSE) +
  geom_errorbar(aes(ymin = lower95, ymax = upper95),colour = 'black',width = 0.05,linewidth = 0.01) +
  geom_point(aes(y = mean_propn_expected),alpha = 0.7, shape = 18) +
  labs(x = 'System type',
       y = '') +
  ylim(c(0,1)) +
  theme(legend.position = 'none')
```


Combine the plots with `patchwork`.

```{r}
acc_type_plts <- all_accuracy_plt + type_acc_plt + plot_layout(widths = c(1, 4))

acc_type_plts

# ggsave('plots/type_accuracy.pdf', acc_type_plts, width = 5.3, height = 2.5, device = cairo_pdf)
```

### Figure 12. Distribution of symmetric conditional entropy z-scores for each system type 

```{r}
systems <- c('0' = 'Type I',
             '1' = 'Type II<sub>side</sub>',
             '2' = 'Type II<sub>gen</sub>',
             '3' = 'Type II<sub>opp</sub>',
             '4' = 'Type III<sub>fb</sub>',
             '5' = 'Type III<sub>mb</sub>',
             '6' = 'Type III<sub>fz</sub>',
             '7' = 'Type III<sub>mz</sub>')


ppt_z %>% group_by(system) %>% mutate(insignificant = ifelse(z < 1.96 & z > -1.96,z,NA), significant = ifelse(z > 1.96 | z < -1.96,z,NA), max = ifelse(true_value == min(true_value), z, NA)) %>% 
  ggplot() +
  geom_histogram(aes(x = insignificant), fill = 'grey',colour = '#666',size = 0.3,bins = 30) +
  geom_histogram(aes(x = significant), fill = '#ddafa9', colour = '#d95f50', size = 0.3,bins = 30) +
  scale_x_continuous(breaks = c(-4,-1.96,0,1.96,4)) +
  labs(x = 'Symmetric conditional entropy (z-transformed)') +
    geom_vline(xintercept = 0,linetype = 'dashed',size = 0.2) +
  facet_wrap(~system, labeller = as_labeller(systems),nrow = 2) +
  theme(
    legend.position = 'none',
    strip.text = ggtext::element_markdown())

# ggsave('plots/participant_z.pdf',height = 3, width = 5.25, device = cairo_pdf)
```


Calculate percentages of significant symmetric conditional entropy.

```{r}
#37.2%
count(ppt_z %>% filter(z < -1.96)) / count(ppt_z) # percentage of participants with significantly low symmetric conditional entropy
```

### Figure 14.


```{r}
all_distance <- rbind(distance,perf_distance)
```

```{r}
# set facet labels
facet_names <- c(
  '1' = 'Type~I',
'2' = 'Type~II[side]',
'3' = 'Type~II[gen]',
'4' = 'Type~II[opp]',
'5' = 'Type~III[fb]',
'6' = 'Type~III[fz]',
'7' = 'Type~III[mb]',
'8' = 'Type~III[mz]',
'9' = 'Type~I',
'10' = 'Type~II[side]',
'11' = 'Type~II[gen]',
'12' = 'Type~II[opp]',
'13' = 'Type~III[fb]',
'14' = 'Type~III[fz]',
'15' = 'Type~III[mb]',
'16' = 'Type~III[mz]'
)
```



```{r}
colours = c('navy','darkred','navy','darkred','navy','darkred','navy','darkred','navy','darkred')

all_distance <- all_distance %>% mutate(
  system = factor(system, levels = c('1','2','3','4','5','6','7','8','9','10','11','12','13','14','15','16')))

heatmaps <- all_distance %>%
  ggplot(aes(x = x, y = y, fill = z)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="black") +
  labs(x = '', y = '') +
  theme(axis.text.x = element_text(colour = colours,angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(colour = colours),
        legend.position = 'none') +
  facet_wrap(~system,nrow = 2, labeller = as_labeller(facet_names, label_parsed))


heatmaps


# ggsave('plots/all_heatmaps.pdf', heatmaps, width = 8.5,height = 3.1, device = cairo_pdf)
```

```{r}
all_distance
```


### Regression models

Did participants generalise predictively?

```{r}
acc_mdl <- glmer(correct_01 ~ 1 + (1|random_id) + (1|system), family = 'binomial', data = acc)

summary(acc_mdl)
```


Did System Type affect whether participants generalised predictively?

```{r}
acc_type_mdl <- glmer(correct_01 ~ system_type + (1|random_id), family = 'binomial', data = acc)

emmeans(acc_type_mdl, list(pairwise~"system_type"),adjust = 'bonferroni')
```


Did participants generalise more predictively when they had gender distinctions in their input compared to non-gender or no distinctions?

```{r}

acc <- acc %>%
  mutate(
    system_group = case_when(
      system == 0 ~ 'None',
      system %in% c(2,3) ~ 'Gender',
      system %in% c(1,4,5,6,7) ~ 'Non-Gender'
    )
  )

acc <- acc %>% mutate(system_group = factor(system_group, levels = c('None','Gender','Non-Gender')))
              
acc_syst_mdl <- glmer(correct_01 ~ system_group + (1|random_id), family = 'binomial', data = acc)

emmeans(acc_syst_mdl, list(pairwise~"system_group"),adjust = 'bonferroni')

```

Did participants encode more gender distinctions when they had gender distinctions in their input compared to non-gender or no distinctions?

```{r}
# merge gender distinctions data with symmetric conditional entropy data
gender_gen <- gender %>% merge(acc, by = c('random_id','block'))

levels(gender_gen$gender)

gender_syst_mdl <- glmer(gender ~ system_group + (1|random_id) + (1|system_group), family = 'binomial', data = gender_gen)

emmeans(gender_syst_mdl, list(pairwise~"system_group"),adjust = 'bonferroni')
```



Do participants generalise less predictively if they saw gender distinct generalisers?

```{r}
# add a column to `acc` telling us whether the given G0 terms ('generalisers') were gender distinct or not.

boys = c('mbs','mzs','fbs','fzs','b')
girls = c('mbd','mzd','fbd','fzd','z')


gender_gen <- gender_gen %>% mutate(
  generalisers = case_when(
    gen1 %in% boys & gen2 %in% girls ~ 1,
    gen1 %in% girls & gen2 %in% boys ~ 1,
    gen1 %in% boys & gen2 %in% boys  ~ 0,
    gen1 %in% girls & gen2 %in% girls ~ 0
  ),
  generalisers = factor(generalisers))


levels(gender_gen$generalisers)


```

```{r}
acc_gen_mdl <- glmer(correct_01 ~ generalisers + (1|random_id) + (1|system.x), family = 'binomial', data = gender_gen)


summary(acc_gen_mdl)
```


Do participants encode more gender distinctions if they saw gender distinct G0 terms?


```{r}
gender_generalisers_mdl <- glmer(gender ~ generalisers + (1|random_id) + (1|system.x), family = 'binomial', data = gender_gen)

summary(gender_generalisers_mdl)
```



